\textbf{Political Instability and Financial Markets}

### 2017-08-07 doc/paper.tex
Updates to introduction and bibliography

\singlespacing
\begin{abstract}
We examine the relationship between political instability and the daily returns of national stock indices. Financial volatility increases dramatically following (and just before) ``irregular'' regime changes caused by coup d'?tat, assassinations, and resignations. Some of the pre-resignation volatility occurs because of public protests such as riots and popular demonstrations that helped spawn the resignation in the first place. However, these irregular regime changes have disparate effects on the direction of stock returns. Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'?tat attempt result in positive abnormal returns of 20\% or more.
\end{abstract}
\doublespacing

\section{Introduction} \label{sec:Introduction}
Investors, firms and development agencies are all concerned with political risk in emerging markets. Government's can reduce asset returns, firm profits and overall economic performance by implementing unfavorable policies. In the 2011 Multilateral Investment Guarantee Agency (MIGA) \textit{World Investment and Political Risk} report, executives of multinational enterprises (MNEs) ranked political risk as the most important constraint for foreign direct investment (FDI) in developing countries over the next three years. The most pressing risks involved government regulatory policies and failed commitments, although political violence was also a concern. \footnote{The types of political risk of most concern to investors in developing countries (ranked in order of importance) were 1) adverse regulatory changes, 2) breaches of contract, 3) transfers and convertibility restrictions, 4) civil disturbances, 5) non-honoring of government guarantees, 6) expropriation nationalization, 7) terrorism and 8) war.}

A common hypothesis is that political instability discourages investment and reduces economic growth by increasing policy uncertainty. Some cross country empirical evidence is consistent with this hypothesis: political instability, as measured by regime change frequency or political violence, is negatively correlated with investment, GDP growth and financial development in cross country regressions \citep{alesina1996political,alesina1996income,roe2011political}. In particular, \citet{alesina1996political} show that countries have lower growth in years in which there are major government changes such as coups. On the other hand, \citet{londregan1990poverty} finds that coups have no effect on economic growth.

Of course, a change in government might promote growth if the new regime adopts better economic policies or is less corrupt than its predecessors. \citet{alesina1996political} takes this possibility seriously but argues that the negative effects of policy uncertainty dominate the positive effects of coups staged by pro growth factions. Conversely, \citet{londregan1990poverty} present evidence suggestive of a bimodal distribution of coups in which neither the effects of pro-growth or anti-growth coups dominate.

As in \citet{alesina1996political}, We use ``irregular'' regime changes such as coups, assassinations and resignations as an indicator for political instability. The major innovation of this paper is that, in contrast to the aforementioned cross country studies, We analyze the impact of political instability using daily financial data. To the extent that these irregular transfers of power are unpredictable and unrelated events do not bias the results, this allows me to use market expectations to quantify the economic impact of regime changes. The advantage of this event study approach is that it allows me to isolate the effect of irregular regime changes and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.

My main finding is that financial volatility increases dramatically following (and just before) coup d'?tat, assassinations, and resignations. However, these irregular regime changes have disparate effects on the direction of stock returns.  Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'?tat attempt result in positive abnormal returns of 20\% or more. We also present evidence showing that volatility increases during public protests such as riots and popular demonstrations that cause leaders to resign. These findings provide important insights into the expected impact of episodes of political instability on economic performance.

This paper is most closely related to four branches of literature. The first analyzes the economic impact of political instability. This area contains a number of theoretical models in addition to the empirical work already mentioned. These models tend to focus on government policy instruments that affect private rates of return on investment. The general idea is that political instability causes governments to act myopically and adopt inefficient policies. For instance, \citet{svensson1998investment} argues that governments are less likely to invest in the legal system and the protection of property rights when political uncertainty is high. Similarly, \citet{devereux1998political} propose a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation causes the private sector to reduce current investment which reduces future output.

The second strand examines the impact of leaders on political and economic outcomes. \citet{jones2005leaders} identify plausibly exogenous leadership transitions using the deaths of leaders from either natural causes or accidents. Their results indicate that leaders influence policy outcomes (particularly monetary policy) and economic growth. \citet{jones2009hit} find that transitions to democracy are more likely after the assassinations of autocrats. Other work has focused on leaders in the private sector. In particular, \citet{bertrand2003managing} estimate CEO fixed effects and show that private sector leaders have important effects on the profits of firms.

The third uses event studies to assess political phenomenon. A growing body of work uses political events to estimate the effect of political connections on firm value \citep[e.g.][]{fisman2001estimating,faccio2006politically,goldman2009politically}. Other studies in ``forensic economics" have used political events to study illicit transactions such as insider trading \citep{dube2011coups}. \citet{guidolin2007diamonds} used the sudden death of rebel leader, Jonas Savimbi, to show that Angolan diamond mining companies benefited from the Angolan civil war. \citet{dellavigna2010detecting} detect illegal arms trading by identifying events that increase or decrease conflict intensity.

The fourth looks at how the variance of stock market returns reacts to political events. \citet{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citet{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies, Luiz In?cio Lula da Silva, increased the volatility (but not the mean return) of the Brazilian stock market.

>>>>

\singlespacing
\begin{abstract}
We examine the relationship between political instability and daily returns of national stock indices. Financial volatility increases dramatically following (and immediately before) ``irregular'' regime changes caused by coup d'etat, assassinations, and resignations. Some of the pre-resignation volatility occurs due to actions that precede the resignation such as public protests, riots, and popular demonstrations. However, these irregular regime changes have disparate effects on the direction of stock returns. Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup attempt result in positive abnormal returns of 20\% or more.
\end{abstract}
\doublespacing

\section{Introduction} \label{sec:Introduction}
For investors, multinational firms, development agencies, and aid organizations, political risk is a top consideration when making investment decisions in emerging markets. In fact, in the 2013 Multilateral Investment Guarantee Agency (MIGA) \textit{World Investment and Political Risk} report, executives of multinational enterprises (MNEs) ranked political risk as the second most important constraint for foreign direct investment (FDI) in developing countries over the next three years (after macroeconomic instability) \citet{wipr2013}. In 2011 and 2012, political risk was ranked the most important constraint for FDI in developing countries - even greater than macroeconomic instability. \footnote{The types of political risk of most concern to investors in developing countries (ranked in order of importance) were 1) adverse regulatory changes, 2) breaches of contract, 3) transfers and convertibility restrictions, 4) civil disturbances, 5) non-honoring of government guarantees, 6) expropriation nationalization, 7) terrorism and 8) war.}

A common hypothesis is that political instability discourages investment and reduces economic growth by increasing policy uncertainty. This hypothesis is supported by cross country empirical evidence that political instability - as measured by regime change frequency or political violence - is negatively correlated with investment, GDP growth, and financial development in cross country regressions \citep{alesina1996political,alesina1996income,roe2011political}. \footnote{In particular, \citet{alesina1996political} show that countries have lower growth in years in which there are major government changes such as coups. On the other hand, \citet{londregan1990poverty} finds that coups have no effect on economic growth.} However, such hypotheses are overly simplistic. Not all political instability is equivalent - resignations, assassinations, and coups are each examples of political instability, but may have disparate effects on financial returns. Further, political instability may be seen as a \textit{positive} event if the current regime is regarded as strongly anti-business or anti-global.  \citet{alesina1996political} takes the possibility of a new government adopting better economic policies seriously, but argues that the negative effects of uncertainty dominate the positive effects of coups staged by pro growth factions.

As in \citet{alesina1996political}, We use ``irregular'' regime changes such as coups, assassinations and resignations as an indicator for political instability. The major innovation of this paper is that, in contrast to the aforementioned cross country studies, we analyze the impact of political instability using daily financial data. This enables us to use market expectations to quantify the economic impact of each individual type of regime change. The advantage of this event study approach is that it allows us to isolate the effect of irregular regime changes and mitigate the endogeneity problems that are nearly ubiquitous in cross country regressions.

Our main finding is that financial volatility increases dramatically following (and just before) coup d'etat, assassinations, and resignations. However, these irregular regime changes have disparate effects on the direction of stock returns. Abnormal returns following resignations are large and positive (6\%) while abnormal returns following assassinations are negative and smaller in magnitude (1\%). The impact of coups tends to be negative, but some events such as the temporary ousting of Hugo Chavez during the 2002 Venezuelan coup d'etat attempt result in positive abnormal returns of 20\% or more. This finding is consistent with research that shows that ``good coups'' - coups that lead to democratization and economic liberalization - are not the norm and therefore coups should not be expected to lead to increased economic development and therefore positive market returns (cite). However, while ``good coups'' may be the minority, they do exist, and coups that attempt to overthrow anti-business autocrats may be expected to result in positive abnormal returns as investors see little risk of a replacement government ``worse'' than the status quo.

This paper is closely related to four branches of literature. The first analyzes the economic impact of political instability. This area contains a number of theoretical models in addition to the empirical work already mentioned. These models tend to focus on government policy instruments that affect private rates of return on investment. The general argument is that political instability causes governments to act myopically and adopt inefficient policies. \footnote{For instance, \citet{svensson1998investment} argues that governments are less likely to invest in the legal system and the protection of property rights when political uncertainty is high. Similarly, \citet{devereux1998political} propose a model in which political instability causes governments to leave fewer assets to their successors which forces them into increasing capital taxes. The knowledge of future taxation causes the private sector to reduce current investment which reduces future output.}

The second looks at how the variance of stock market returns reacts to political events. \citet{leblang2005government} show that right-wing governments increase the mean and volatility of stock prices in the United States and Britain. \citet{jensen2005market} find that a rise in the popularity of a Brazilian presidential candidate with uncertain future policies increased the volatility (but not the mean return) of the Brazilian stock market.

The third strand examines the impact of leaders on political and economic outcomes. \citet{jones2005leaders} identify exogenous leadership transitions using deaths of leaders from natural causes or accidents. Their results indicate that leaders influence monetary policy and economic growth. \citet{jones2009hit} find that transitions to democracy are more likely after the assassinations of autocrats.

The fourth uses event studies to assess political phenomenon. A growing body of work uses political events to estimate the effect of political connections on firm value \citep[e.g.][]{fisman2001estimating,faccio2006politically,goldman2009politically}. Studies in ``forensic economics" have used political events to examine illicit transactions such as insider trading \citep{dube2011coups}. \citet{guidolin2007diamonds} used the sudden death of rebel leader to show that Angolan diamond mining companies benefited from the Angolan civil war. \citet{dellavigna2010detecting} detect illegal arms trading by identifying events that increase or decrease conflict intensity.
